gosset: An R package for analysis and synthesis of ranking data in agricultural

### 2022-06-20 inst/paper/paper.Rmd
update manuscript

---
title: 'gosset: An R package for analysis and synthesis of ranking data in agricultural
  experimentation'
abstract: null
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

>Kauê de Sousa^1,2[*]^, Jacob van Etten^2^, David Brown^3,4^, Jonathan Steinke^2,5^  
>-
>
^1^ Department of Agricultural Sciences, Inland Norway University of Applied Sciences, 2318 Hamar, Norway  
^2^ Digital Inclusion, Bioversity International, Parc Scientifique Agropolis II, 34397, Montpellier Cedex 5, France   
^3^ Laboratory of Geo-Information Science and Remote Sensing, Wageningen University & Research, Droevendaalsesteeg 3, 6708 PB, Wageningen, The Netherlands   
^4^ Digital Inclusion, Bioversity International, 30501, Turrialba, Costa Rica   
^5^ Humboldt University Berlin, Berlin, Germany

@@

::: frontmatter
::: keyword
data-driven agriculture, Plackett-Luce model
:::
:::

@@

The R package gosset provides functionality supporting the analysis
workflows in agricultural experimentation, especially with rank-based
approaches. The package is available in The Comprehensive R Archive Network (CRAN)
and can be installed by executing ``install.packages("gosset")``.

@@

When data from agricultural experiments is not in ranking format, it
should be transformed to be used as inputs into R packages for the
analysis of ranking data. For instance, the Plackett-Luce model [@luce_individual_1959;@Plackett]
is implemented in the R package as PlackettLuce,
which requires the data to be formatted as ranking matrix. Another
example is the Bradley-Terry model [@bradley1952rank], implemented
in the package BradleyTerry2 [@BradleyTerry2] and which requires the
input data to be formatted as paired comparisons. For these cases,
gosset provides the functions ``rank_numeric`` and ``rank_binomial`` respectively. The
function ``rank_numeric`` transforms a set of numeric values into an ordinal
ranking, considering if items with higher numeric values should be ranked first or
last. The function ``rank_binomial`` transforms data in rankings format into
pairwise comparisons, as required by the package BradleyTerry2.
Additionally, gosset provides the function ``rank_tricot``, for the case
when the experimental data is collected from tricot trials. In those trials,
farmers rank the technology under evaluation (e.g., crop variety) stating which
is the best and which is the worst from a set of three [@vanetten_beza_2019].

@@

The gosset package provides the following functions to support model
selection and validation of Bradley-Terry and Plackett-Luce models. The function ``pseudoR2``
computes goodness-of-fit measure McFadden's pseudo-R2 [@mcfadden1973conditional]. The ``AIC``
function computes the Akaike Information Criterion[@akaike1974]. The ``kendallTau`` function
computes the Kendall-tau rank correlation coefficient [@kendall_1938].

@@

``plot``

``worth_map``

``worth_map``

``regret``

# Illustrative Examples
We demonstrate the functionality of the gosset package using the 
trial dataset "nicabean", which consists of trial data collected in
Nicaragua following the tricot approach. We use the Plackett-Luce model
implemented in the R package *PlackettLuce*[@Turner2020]. We use climate data
as model covariates to investigate the effect fo climate factors on the performance 
of common bean (*Phaseolus vulgaris* L.) genotypes. For obtaining the climate data,
we use the *nasapower* package[@nasapower]. Climatic indices were computed with the
*climatrends* package[@climatrendspkg].

First, the required packages and data are loaded.
``` {r}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("nasapower")

#data("nicabean", package = "gosset")

load("D:/GitHub/gosset/data/nicabean.rda")

@@

```
Make a PlackettLuce rank using the function ``rank_numeric``
```{r}
R <- vector(mode = "list", length = length(traits))

@@

```{r}
head(R[[1]], 10)
```
Kendall correlation between traits using kendallTau()
```

>>>>

---
title: 'gosset: An R package for analysis and synthesis of ranking data in agricultural
  experimentation'
abstract: null
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

>Kauê de Sousa^1,2[*]^, David Brown^3,4^, Jonathan Steinke^2,5^, Jacob van Etten^2^  
>-
>
^1^ Department of Agricultural Sciences, Inland Norway University of Applied Sciences, 2318 Hamar, Norway  
^2^ Digital Inclusion, Bioversity International, Parc Scientifique Agropolis II, 34397, Montpellier Cedex 5, France   
^3^ Laboratory of Geo-Information Science and Remote Sensing, Wageningen University & Research, Droevendaalsesteeg 3, 6708 PB, Wageningen, The Netherlands   
^4^ Digital Inclusion, Bioversity International, 30501, Turrialba, Costa Rica   
^5^ Humboldt University Berlin, Berlin, Germany

@@

::: frontmatter
::: keyword
**Keywords**: Bradley-Terry, Plackett-Luce, data science, on-farm trials, participatory-research, tricot
:::
:::

@@

The R package gosset provides functionality supporting the analysis
workflows in agricultural experimentation, especially with rank-based
approaches. The package is available in The Comprehensive R Archive Network (CRAN)
and can be installed by executing ``install.packages("gosset")``. The package is named 
in honor to William Sealy Gosset, known by the pen name ‘Student’, a pioneer of modern statistics
in small sample experimental design and analysis [@Ziliak_2019].

@@

*   **``rank_binomial``** transforms a ranking object into a binary comparisons, as required
by package BradleyTerry2 [@BradleyTerry2].  
*   **``rank_numeric``** converts numeric values into rankings. The parameter ``ascending``
indicates if the rankings should be made considering the numeric values in ascending order. 
The default is ``asceding = FALSE``. This function is useful when the data were collected 
as numerical observations, for instance, in a experiment measuring crop yield.  
*   **``rank_tricot``** transforms data in tricot format into rankings.

@@

*   **``AIC``** computes the Akaike Information Criterion @akaike1974] for a Bradley-Terry model fitted with
BradleyTerry2 [@BradleyTerry2] or a Plackett-Luce model fitted with [@Turner2020].  
*   **``btpermute``** deviance-based forward variable selection [@lysen2009permuted] procedure for Bradley-Terry models.  
*   **``compare``** compare the agreenment between methods.  
*   **``crossvalidation``** performs k-fold cross-validation, where k could be specified by the user.The default is
10-fold. Folds can be provided as a vector for a custom cross-validation, such as blocked cross-validation.
*   **``forward``** executes forward variable selection with cross-validation.  
*   **``kendallTau``** computes the Kendall-tau rank correlation coefficient between two rankings [@kendall_1938].  
*   **``kendallW``** computes Kendall's W (coefficient of concordance) among observed rankings and those predicted by
the Plackett-Luce model.  
*   **``pseudoR2``** computes goodness-of-fit measure McFadden's pseudo-R2 [@mcfadden1973conditional].

@@

*   **``plot``** provides an alternative to the default plot function of the PlackettLuce package.

*   **``regret``**

*   **``reliability``**

*   **``worth_bar``**

*   **``worth_map``**

# Illustrative Examples
We demonstrate the functionality of the gosset package using the 
trial dataset "nicabean", which consists of trial data collected in
Nicaragua. In this example we present a suggested workflow to assess trait prioritization and crop performance using decentralized on-farm data generated with the tricot approach [@vanetten_beza_2019].

We use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot approach, farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties has the best and worst performance in nine traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste) and overall appreciation). The farmers also provided their overall appreciation about the varieties, i.e., which
variety has the best and the worst performance based on the overall performance considering all the traits.

We use the Plackett-Luce model implemented in the R package *PlackettLuce*[@Turner2020]. 
To consider the effecto of climate on the variety performance, we use climate data
For obtaining the climate data, we used the *nasapower* package[@nasapower]. 
Climatic indices were computed with the *climatrends* package[@climatrendspkg].

First, we load the required packages and data. The ``nicabean`` dataset contains the 
tricot rankings (``trial``), the climate covariates (``covar``) and the traits observations (``traits``).
``` {r fetch, message=FALSE, eval=TRUE, echo=TRUE}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("nasapower")

data("nicabean", package = "gosset")

@@

```
Make a PlackettLuce rank using the function ``rank_numeric``.
```{r}
R <- vector(mode = "list", length = length(traits))

@@

```{r}
head(R[[1]], 5)
```

Kendall correlation between traits using kendallTau()
```
